---
title: "VCE Based On Subsampled Data"
author: "Peter von Rohr"
date: "2019-01-08"
output: html_notebook
---


## Disclaimer
Different strategies of sub-sampling data from a large data set are evaluated with respect to variance components estimation.


## Introduction
Breeding values are predicted using BLUP animal models (BLUP-AM). BLUP-AM are specialized members of the larger class of linear mixed effect models (LME). In BLUP-AM, predicted breeding values (and estimates for fixed effects) are obtained by solving mixed-model equations which require known variance components from all random components. 


## Variance Components Estimation (VCE)
Breeding values are predicted for very large data sets. Normally, these data sets are too large to be used in variance components estimation. Hence we need a method to draw random samples from the data to get estimates of variance components. The goal is to use as small a data sample as possible to still get reliable estimates of variance components.


## Numerical Example
We want to use the data sets in `pedigreemm` to do a few experiments with data sub-sampling. In a first step, we are just reproducing the first example shown in Vazquez et al. (2010). This leads to the following analysis

```{r}
require(pedigreemm)
milk <- within(milk, sdMilk <- milk / sd(milk))

### # either run the analysis, or load the results from the file
s_result_file <- "fm1.rda"
if (file.exists(s_result_file)){
  load(s_result_file)
} else {
system.time(
  fm1 <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd), 
                    data = milk, pedigree = list(id = pedCowsR)))
  save(fm1, file = s_result_file)  
}
```

The above output gives us the timing of the analysis. Now we want to have a look at the results of the analysis.

```{r}
summary(fm1)
```


## Subsampling
In our example the complete data set contains `r nrow(milk)` records. From the complete record, we want to take a random sample of a given size and from this smaller data set variance components should be estimated again.


## Strategies
Several strategies are possible to generate smaller data sets from the complete data set. The following list presents some of them

* Randomly select records from the large data set
* Keep the records coming from the largest herds


### Random Selection
The strategy of `random selection` mean that we are just randomly choosing observations out of the complete data set. This is done via the function `sample()` which draws a random sample out of a large data set.

```{r}
set.seed(432)
n_sample_size <- 500
vec_rsample_idx <- sample(1:nrow(milk), size = n_sample_size)
df_rsample_milk <- milk[vec_rsample_idx,]
dim(df_rsample_milk)
head(df_rsample_milk)
```

Now the same analysis is done using the smaller sample data set.

```{r}
s_result_file_rsample <- "fm1rsample.rda"
if (file.exists(s_result_file_rsample)){
  load(s_result_file_rsample)
} else {
system.time(
  fm1rsample <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd), 
                    data = df_rsample_milk, pedigree = list(id = pedCowsR)))
  save(fm1, file = s_result_file_rsample)  
}
```

The results are

```{r}
summary(fm1rsample)
```


### Largest Herds
Whether the sub-sampling based on herd size is a good strategy or not could depend on the distribution of herd sizes. The following frequency table and the following histogram gives a first impression on the distribution

```{r}
table(milk$herd)
```

```{r}
hist(as.numeric(milk$herd))
```


Drawing the histogram using `ggplot2` leads to 

```{r}
require(ggplot2)
ggplot(milk, aes(herd)) + geom_histogram(stat = "count")
```


