---
title: "VCE Based On Subsampled Data"
author: "Peter von Rohr"
date: "2019-01-08"
output: html_notebook
---


## Disclaimer
Different strategies of sub-sampling data from a large data set are evaluated with respect to variance components estimation.


## Introduction
Breeding values are predicted using BLUP animal models (BLUP-AM). BLUP-AM are specialized members of the larger class of linear mixed effect models (LME). In BLUP-AM, predicted breeding values (and estimates for fixed effects) are obtained by solving mixed-model equations which require known variance components from all random components. 


## Variance Components Estimation (VCE)
Breeding values are predicted for very large data sets. Normally, these data sets are too large to be used in variance components estimation. Hence we need a method to draw random samples from the data to get estimates of variance components. The goal is to use as small a data sample as possible to still get reliable estimates of variance components.


## Numerical Example
We want to use the data sets in `pedigreemm` to do a few experiments with data sub-sampling. In a first step, we are just reproducing the first example shown in Vazquez et al. (2010). This leads to the following analysis

```{r}
require(pedigreemm)
milk <- within(milk, sdMilk <- milk / sd(milk))

### # either run the analysis, or load the results from the file
s_result_file <- "fm1.rda"
if (file.exists(s_result_file)){
  load(s_result_file)
} else {
system.time(
  fm1 <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd), 
                    data = milk, pedigree = list(id = pedCowsR)))
  save(fm1, file = s_result_file)  
}
```

The above output gives us the timing of the analysis. Now we want to have a look at the results of the analysis.

```{r}
summary(fm1)
```


## Subsampling
In our example the complete data set contains `r nrow(milk)` records. From the complete record, we want to take a random sample of a given size and from this smaller data set variance components should be estimated again.


## Strategies
Several strategies are possible to generate smaller data sets from the complete data set. The following list presents some of them

* Randomly select records from the large data set
* Keep the records coming from the largest herds


### Random Selection
The strategy of `random selection` mean that we are just randomly choosing observations out of the complete data set. This is done via the function `sample()` which draws a random sample out of a large data set.

```{r}
set.seed(432)
n_sample_size <- 500
vec_rsample_idx <- sample(1:nrow(milk), size = n_sample_size)
df_rsample_milk <- milk[vec_rsample_idx,]
dim(df_rsample_milk)
head(df_rsample_milk)
```

Now the same analysis is done using the smaller sample data set.

```{r}
s_result_file_rsample <- "fm1rsample.rda"
if (file.exists(s_result_file_rsample)){
  load(s_result_file_rsample)
} else {
system.time(
  fm1rsample <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd), 
                    data = df_rsample_milk, pedigree = list(id = pedCowsR)))
  save(fm1rsample, file = s_result_file_rsample)  
}
```

The results are

```{r}
summary(fm1rsample)
```

The same subsampling of the data should be repeated and we want to collect the variance estimates. First, we have to check how to extract the variance estimates from `fm1rsample`. The function `VarCorr()` can be used to extract the variance components. 

```{r}
?VarCorr
```


```{r}
o_cur_rsample <- VarCorr(fm1rsample)
print(o_cur_rsample, comp=c("Variance","Std.Dev."))
```

The single components are obtained by 

```{r}
o_cur_rsample$id[1,1]
```

for the genetic additive variance, by

```{r}
o_cur_rsample$herd[1,1]
```

for the herd variance and by 

```{r}
str(o_cur_rsample)
```

```{r}
str(fm1rsample)
```


```
FIXME: The following statement does not return the residual variance component estimate but its standard deviation
```


```{r}
attr(o_cur_rsample, "sc")
```

for the residual variance. Now we are ready to repeat the sub-sampling a number of times and get boot-strapped results. 

```{r}
n_nr_rep <- 3
l_rep_results <- sapply(1:n_nr_rep, 
                        function(x, pn_sample_size){
                          vec_rsample_idx <- sample(1:nrow(milk), size = n_sample_size)
                          df_rsample_milk <- milk[vec_rsample_idx,]
                          fm1rsample <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd), 
                                                   data = df_rsample_milk, pedigree = list(id = pedCowsR))
                          o_cur_rsample_result <- VarCorr(fm1rsample)
                          return(list(id = o_cur_rsample_result$id[1,1],
                                      herd = o_cur_rsample_result$herd[1,1],
                                      residual = attr(o_cur_rsample_result, "sc")))
                        }, n_sample_size)
```

From these results, we take the mean and the standard deviation

```{r, echo=FALSE, results='asis'}
vec_components <- c("Id", "Herd", "Residual")
vec_rsample_mean <- unlist(sapply(vec_components, function(x) mean(unlist(l_rep_results[tolower(x),]))))
vec_rsample_sd <- unlist(sapply(vec_components, function(x) sd(unlist(l_rep_results[tolower(x),]))))
tbl_rsample_result <- tibble::data_frame(Component = vec_components,
                                         `Sub Sample Mean` = vec_rsample_mean,
                                         `Sub Sample Sd` = vec_rsample_sd)
knitr::kable(tbl_rsample_result)
```


### Largest Herds
Whether the sub-sampling based on herd size is a good strategy or not could depend on the distribution of herd sizes. The following frequency table and the following histogram gives a first impression on the distribution

```{r}
table(milk$herd)
```

```{r}
hist(as.numeric(milk$herd))
```


Drawing the histogram using `ggplot2` leads to 

```{r}
require(ggplot2)
ggplot(milk, aes(herd)) + geom_histogram(stat = "count")
```


When data are to be sampled based on herd size, we have to sort the original data set according to herdsize. 

```{r}
require(dplyr)
tbl_milk <- tibble::as_tibble(milk)
tbl_milk_hs <- tbl_milk %>% 
  group_by(herd) %>%
  summarize(herdsize = n()) %>%
  arrange(desc(herdsize)) %>%
  mutate(cumhs = cumsum(herdsize))

tbl_milk_hs
```

To get a certain sample size, a filter on the cumulative herd size is applied.

```{r}
tbl_hs_herd <- tbl_milk_hs %>% filter(cumhs < n_sample_size)
tbl_hs_herd
```

Because, we want to have a dataset with at least `r n_sample_size` of observations, we include an additional row from the sorted `tbl_df`. 

```{r}
n_last_row_chosen <- nrow(tbl_hs_herd)
tbl_hs_herd <- tbl_mil_hs_sample %>% bind_rows(tbl_milk_hs[(n_last_row_chosen+1),])
tbl_hs_herd
```

The tibble `tbl_hs_herd` is used to select observations from the original dataset by joining `tbl_hs_herd` back to the original `tbl_milk` using the column `herd` as joining criterion.

```{r}
tbl_milk_hs_sample <- tbl_milk %>% inner_join(tbl_hs_herd)
tbl_milk_hs_sample
```

This sample tibble has dimensions

```{r}
dim(tbl_milk_hs_sample)
```

Now the analysis is done based on this sample tibble dataframe

```{r}
s_hs_sample <- "fm1hssample.rda"
if (file.exists(s_hs_sample)){
  load(s_hs_sample)
} else {
  fm1hssample <- pedigreemm(sdMilk ~ lact + log(dim) + (1|id) + (1|herd), 
                    data = tbl_milk_hs_sample, pedigree = list(id = pedCowsR))
  save(fm1hssample, file = s_hs_sample)
  
}
summary(fm1hssample)
```

These results show a clearly reduced herd-variance component and an increased value of the residual variance component compared to the estimate of the complete dataset. The estimate of the genetic-additive variance is comparable to what was obtained with the complete dataset. Selecting a subset of the data just on the criterion of large herd sizes does not give us the opportunity to repeat the vce based on different subset selections as was the case in the random selection strategy. This can be compensated by setting a minimum number of observation per herd and then randomly select from this smaller dataset.


## Connectedness
```
TBC
```

